---
output: html_document
bibliography: references.bib
csl: bioinformatics.csl
link-citations: yes
author: 
  - Matteo Calgaro^[University of Verona, Verona, Italy, matteo.calgaro_01@univr.it]
  - Andrea Telatin^[Quadram Institute Bioscience, Norwich, UK, andrea.telatin@quadram.ac.uk]
---

# Rarefaction Practice in Microbiome Data Analysis

In the last decade, there has been ongoing debate regarding the optimal approach for normalizing sequencing depth in samples analyzed using 16S rRNA gene sequencing. A highly influential paper by McMurdie and Holmes [@mcmurdie2014] strongly argued against the practice of rarefying sequence counts, deeming it "inadmissible" and discouraging its utilization. Despite a rebuttal on this subject [@weiss2017], which demonstrated the benefits of rarefying in certain cases, the proponents of alternative normalization method seem to have a predominant influence within the microbiome research community. Nevertheless, a recent publication by Patrick Schloss [@schloss2023], shows that rarefaction is the most robust approach to control for uneven sequencing effort when considered across a variety of alpha and beta diversity metrics.

## Aims of this chapter

-   Collect some definitions and clarify the difference between the terms *"rarefy"* and *"rarefaction"*;

-   exemplify the previous terms in alpha and beta diversity analyses;

-   provide the code to implement rarefaction curves and to estimate rarefaction estimates of richness and Bray-Curtis dissimilarities.

## Required packages

List of libraries used:

-   [mia](https://github.com/microbiome/mia#readme) (`BiocManager::install("mia")`)
-   [miaViz](https://microbiome.github.io/miaViz/) (`BiocManager::install("miaViz")`)
-   [scater](https://bioconductor.org/packages/release/bioc/html/scater.html) (`scaterBiocManager::install("scater")`)
-   [patchwork](https://patchwork.data-imaginist.com/) (`install.packages('patchwork')`)
-   [ranacapa](https://github.com/gauravsk/ranacapa) (`devtools::install_github("gauravsk/ranacapa")`)
-   [BiocParallel](https://bioconductor.org/packages/release/bioc/html/BiocParallel.html) (`BiocManager::install("BiocParallel")`)

## Introduction and definitions

Rarefaction in microbiome research was introduced from ecology research out of the need to address the issue of uneven sampling depth. When studying microbial communities using high-throughput sequencing techniques like 16S rRNA gene sequencing or shotgun metagenomics, the number of sequencing reads obtained from different samples can vary significantly. This uneven sampling depth can introduce biases and make it challenging to compare the diversity and abundance of microbial taxa between samples accurately.

Rarefaction standardises the sampling effort across all samples and should ensure fair comparisons of diversity and abundance. By repeatedly subsampling the sequencing data to a common sequencing depth for each sample, the rarefaction method allows researchers to obtain a consistent and unbiased estimate of the microbial diversity present in each sample.

Library size normalization by random subsampling without replacement is called *rarefying*. There is confusion in the literature regarding terminology, and sometimes this normalization approach is conflated with a non-parametric resampling technique, called *rarefaction*. According to [@mcmurdie2014], rarefying is most often defined by the following steps:

1.  Select a minimum library size, $N_{L,min}$. This has also been called the rarefaction level.

2.  Discard libraries (microbiome samples) that have fewer reads than $N_{L,min}$.

3.  Subsample the remaining libraries without replacement such that they all have size $N_{L,min}$.

Let's make a practical example in R. We start by creating a very simple count matrix with 3 taxa and 2 samples.

```{r}
# Generate a very simple count matrix where the first
# sample has half the library size of the second sample
count_matrix <- matrix(c(68, 32, 200, 200, 200, 200), 
    nrow = 3, ncol = 2)
colnames(count_matrix) <- c("S1", "S2")
rownames(count_matrix) <- c("Taxa1", "Taxa2", "Taxa3")

count_matrix
```

`S2` sample has a library size which is twice the library size of sample `S1`.

```{r}
colSums(count_matrix)
```

According to the definition, we choose $N_{L,min} = 300$ and we subsample, without replacement (random number generation seed equal to 123, see [why](https://www.tutorialspoint.com/why-we-should-use-set-seed-in-r)), the counts such that they all the samples have size $300$.

```{r}
# Function to rarefy counts of a count matrix
rarefy_function <- function(count_matrix) {
  
  # Determine the minimum library size across samples
  minLibSize <- min(colSums(count_matrix))
  
  # Perform rarefaction on each sample in the count matrix
  count_rarefy <- apply(count_matrix, 2, FUN = function(smpl){
    
      # Create a vector of taxon names proportional to their counts in the sample
      vec <- rep(rownames(count_matrix), times = smpl)
        
      # Perform subsampling by randomly selecting taxon names without replacement
      subsampling <- sample(vec, size = minLibSize, replace = FALSE)
        
      # Create a table of the subsampled taxon counts
      table(subsampling)
  })
  
  return(count_rarefy)
}

# Set the random seed for reproducibility 
set.seed(123)

rarefied_counts <- rarefy_function(count_matrix)
rarefied_counts
```

The resulting table is a table where the `S1` remains the same as before subsampling. Indeed, subsampling $N_{L,min} = 300$ counts from a sample with that same library size is like taking the entire sample. `S2` instead, has now a total of 300 counts divided across the three taxa.

As the subsampling step is performed only once, another way to obtain the rarefied counts more easily, is to divide the counts by their library size, multiply the relative abundances by $N_{L,min} = 300$, and round the results.

```{r}
rarefy_simple_function <- function(count_matrix) {
  
  # Determine the minimum library size across samples
  minLibSize <- min(colSums(count_matrix))
  
  # Perform simple rarefaction on each sample in the count matrix
  count_rarefy_simple <- apply(count_matrix, 2, FUN = function(smpl){
    
    # Calculate the relative abundance of each taxon in the sample
    rel_abundance <- smpl / sum(smpl)
    
    # Multiply the relative abundances by Lmin to obtain the rarefied counts
    rarefied_counts <- round(minLibSize * rel_abundance, digits = 0)
    
    # Return the rarefied counts for the current sample
    return(rarefied_counts)
  })
  
  return(count_rarefy_simple)
}
count_rarefied_simple <- rarefy_simple_function(count_matrix)
count_rarefied_simple
```

According to [@schloss2023], the authors of [@mcmurdie2014] were correct to state that the distinction between "rarefying" and "rarefaction" was confusing and led to their conflation. However, they poorly managed to solve the problem due to misleading sentences throughout the publication. Traditionally, repeating the subsampling step a large number of times and averaging the result is called *rarefaction*. Instead, *rarefying* or *subsampling* is rarefaction, but with a single randomization. To minimize confusion, we will use "subsampling" in place of "rarefying" through the rest of this chapter. For clarity, we will use the same definition of rarefaction of [@schloss2023]:

1.  Select a minimum library size, $N_{L,min}$. Researchers are encouraged to report the value of $N_{L,min}$.

2.  Discard samples that have fewer reads than $N_{L,min}$.

3.  Subsample the remaining libraries without replacement such that they all have size $N_{L,min}$.

4.  Compute the desired metric (*e.g.*, richness, Shannon diversity, Bray-Curtis distances) using the subsampled data.

5.  Repeat steps 3 and 4 a large number of iterations (typically 100 or 1,000). Researchers are encouraged to report the number of iterations.

6.  Compute summary statistics (*e.g.*, the mean) using the values from step 4.

## Rarefaction in practice

The *rarefaction*, *i.e.* repeating the subsampling step a large number of times and averaging the result, is here exemplified for [Richness index](https://www.coastalwiki.org/wiki/Measurements_of_biodiversity) and [Bray-Curtis dissimilarity index](https://en.wikipedia.org/wiki/Bray%E2%80%93Curtis_dissimilarity).

Let's start by using an example dataset from a research published in PNAS in early 2011 [@caporaso2011]. This work compared the microbial communities from 25 environmental samples and three known "mock communities" -- a total of 9 sample types -- at a depth averaging 3.1 million reads per sample. The [dataset](https://microbiome.github.io/mia/reference/mia-datasets.html) comes with the [mia](https://microbiome.github.io/mia/) package.

```{r, message=FALSE, warning=FALSE}
library(mia)
data("GlobalPatterns", package = "mia")

# See https://microbiome.github.io/mia/reference/agglomerate-methods.html
gp_genus <- mia::agglomerateByRank(GlobalPatterns, "Genus")

# Filter out rows (taxa) with zero counts across all samples
gp_genus <- gp_genus[
    rowSums(assay(gp_genus, "counts")) > 0, ]

library(scater)
gp_genus <- addPerCellQC(gp_genus)
```

Here we show the distribution of library sizes across samples. We have library sizes ranging from 58688 reads to 40x times more.

```{r, fig.cap="Sample library sizes grouped and coloured by sample type."}
library(miaViz)
summary(colData(gp_genus)$sum)

# Create a scatter plot of the 'gp_genus' dataset, where the y-axis represents the 'sum' column in the colData,
# the x-axis represents the 'SampleType' column in the colData, and the points are colored by the 'SampleType' column
plotColData(object = gp_genus, 
    y = "sum", x = "SampleType", 
    colour_by = "SampleType") + 
    theme(axis.text.x = element_text(angle = 45, hjust=1))
```

### Richness index

Richness refers to the overall count of species within a community (sample). The most basic richness index corresponds to the number of observed species (observed richness). Richness estimates remain unaffected by the abundances of individual species.

To compute richness we use the [`estimateRichness`](https://rdrr.io/github/microbiome/mia/man/estimateRichness.html) function of the `mia` package.

```{r}
gp_genus <- mia::estimateRichness(gp_genus, 
    assay.type = "counts", 
    index = "observed")
```

We can inspect the observed richness distribution of each sample type:

```{r, fig.cap="Richness estimates grouped and coloured by sample type."}
p_r <- plotColData(object = gp_genus, 
    y = "observed", x = "SampleType", 
    colour_by = "SampleType") + 
    theme(axis.text.x = element_text(angle = 45, hjust=1)) + 
    labs(title = "Richness in raw counts")
p_r
```

And we can also describe the relationship between library size and richness. When the sequencing depth is enough to describe the samples we do not expect to see any particular pattern in this kind of graphical representation.

```{r, fig.cap="Relationship between richness and library size coloured by sample type."}
plotColData(object = gp_genus, 
    x = "sum", y = "observed", 
    colour_by = "SampleType") + 
    geom_smooth() + 
    labs(x = "Library Size", y = "Richness", 
         title = "Richness vs Library Size")
```

But all of the above plots are generated from raw data. Here we want to inspect the effect of rarefaction. For this reason we use the [`subsampleCounts`](https://rdrr.io/github/FelixErnst/mia/man/subsampleCounts.html) function of the package `mia` to obtain subsampled counts. We repeat this process 3 times with 3 different seeds (1, 2, and 3). We store the results in three differentially named assays of the `TreeSummarizedExperiment` object: *rare1*, *rare2*, and *rare3*.

```{r, warning=FALSE, message=FALSE}
gp_genus <- mia::subsampleCounts(gp_genus, 
    assay.type = "counts", 
    min_size = min(gp_genus$sum), 
    seed = 1, replace = FALSE, 
    verbose = FALSE, 
    name = "rare1")
gp_genus <- mia::subsampleCounts(gp_genus, 
    assay.type = "counts", 
    min_size = min(gp_genus$sum), 
    seed = 2, replace = FALSE, 
    verbose = FALSE, 
    name = "rare2")
gp_genus <- mia::subsampleCounts(gp_genus, 
    assay.type = "counts", 
    min_size = min(gp_genus$sum), 
    seed = 3, replace = FALSE, 
    verbose = FALSE, 
    name = "rare3")
```

We compute richness for each subsampled count matrix:

```{r}
gp_genus <- estimateRichness(gp_genus, assay.type = "rare1", name = "rich1", index = "observed")
gp_genus <- estimateRichness(gp_genus, assay.type = "rare2", name = "rich2", index = "observed")
gp_genus <- estimateRichness(gp_genus, assay.type = "rare3", name = "rich3", index = "observed")
```

And plot the results:

```{r, fig.width=10, fig.height=8, fig.cap="Richness estimates grouped and coloured by sample type. a) Raw data. b) subsampled counts, seed = 1. c) subsampled counts, seed = 2. d) subsampled counts, seed = 3."}
p_r1 <- plotColData(object = gp_genus, 
    y = "rich1", x = "SampleType", 
    colour_by = "SampleType") + 
    theme(axis.text.x = element_text(angle = 45, hjust=1)) + 
    labs(title = "Subsampled counts - seed = 1", 
         subtitle = paste("N(L,min) =", min(gp_genus$sum)))
p_r2 <- plotColData(object = gp_genus, 
    y = "rich2", x = "SampleType", 
    colour_by = "SampleType") + 
    theme(axis.text.x = element_text(angle = 45, hjust=1)) + 
    labs(title = "Subsampled counts - seed = 2", 
         subtitle = paste("N(L,min) =", min(gp_genus$sum)))
p_r3 <- plotColData(object = gp_genus, 
    y = "rich3", x = "SampleType", 
    colour_by = "SampleType") + 
    theme(axis.text.x = element_text(angle = 45, hjust=1)) + 
    labs(title = "Subsampled counts - seed = 3", 
         subtitle = paste("N(L,min) =", min(gp_genus$sum)))

library(patchwork)
(plot_spacer() + p_r + plot_spacer()) / (p_r1 + p_r2 + p_r3) + 
    plot_layout(guides = "collect") + plot_annotation(tag_levels = "a")
```

The main difference in these panels are the range of the y-axis: it reaches higher values in **a** rather than **b**, **c**, and **d**. Moreover, while the richest samples belonged to Freshwater (creek) when the richness was evaluated using raw counts (panel **a**), Sediment (estuary) samples became the richest when the subsampled counts were used instead (**b**, **c**, and **d** panels). This is probably due to the differential presence of rare taxa between the two environments. Minor differences are observable when comparing **b**, **c**, and **d** panels, but still they are present.

Practically, rarefaction consists in averaging the index measures of panels **b**, **c**, and **d** to obtain a single value for each sample.

```{r, fig.cap="Richness estimates grouped and coloured by sample type. Richness indexes were averaged across the 3 subsamplings to obtain a single value for each sample."}
gp_genus$rich_rarefaction <- rowMeans(
    as.data.frame(colData(gp_genus)[, 
        c("rich1", "rich2", "rich3")]))
p_rarefaction <- plotColData(object = gp_genus, 
    y = "rich_rarefaction", x = "SampleType", 
    colour_by = "SampleType") + 
    theme(axis.text.x = element_text(angle = 45, hjust=1)) + 
    labs(title = "Rarefaction Richness", 
         subtitle = paste("Subsamplings = 3, N(L,min) =", 
            min(gp_genus$sum)))
p_rarefaction
```

To automatise the entire procedure the following function can be used:

```{r, warning=FALSE, message=FALSE}
library(BiocParallel)

rarefaction_richness <- function(
    tse, # The object to use
    assay.type = "counts", # Assay to use
    subsamplings = 100, # Number of subsamplings
    minLibSize = NULL, # The desired library size 
    seed = 123, # The seed
    BPPARAM = SerialParam()){ # Parallelisation
    
    counts <- assay(tse, "counts")
    
    if(is.null(minLibSize))
        minLibSize <- min(colSums(counts))
    
    richness_list <- bplapply(
        X = 1:subsamplings, 
        BPPARAM = BPPARAM, 
        BPOPTIONS = bpoptions(RNGseed = seed), 
        FUN = function(sub){
            # Subsampling step
            tse <- mia::subsampleCounts(tse, 
                assay.type = assay.type, 
                min_size = minLibSize,
                # seed = runif(1, 0, .Machine$integer.max), 
                replace = FALSE, 
                verbose = FALSE, 
                name = "sub")
            # Estimation step
            richness <- mia::estimateRichness(tse, 
                assay.type = "sub", 
                index = "observed", 
                name = "richness")$richness
            return(richness)
    })
    # Return the averaged values
    colMeans(plyr::ldply(richness_list))
}
```

We run it in parallel (Linux or MacOS, Windows users should use `BPPARAM = SerialParam()`) and we directly store the computed values inside the `TreeSummarizedExperiment` object.

```{r}
gp_genus$rarefaction_richness <- rarefaction_richness(
    tse = gp_genus, 
    assay.type = "counts", 
    subsamplings = 100, 
    BPPARAM = MulticoreParam(4), 
    minLibSize = NULL, 
    seed = 123)
```

Finally, the graphical representation of rarefied richness (seed = 123, number of subsamplings = 100, $N_{L,min} = 58688$) is reported below.

```{r, fig.cap="Richness estimates grouped and coloured by sample type. Richness indexes were averaged across 100 subsamplings (rarefaction library size = 58688) to obtain a single value for each sample (seed = 123)."}
p_rarefaction100 <- plotColData(object = gp_genus, 
    y = "rarefaction_richness", x = "SampleType", 
    colour_by = "SampleType") + 
    theme(axis.text.x = element_text(angle = 45, hjust=1)) + 
    labs(title = "Rarefaction Richness", 
         subtitle = paste("Seed = 123, Subsamplings = 100, N(L,min) =", 
            min(gp_genus$sum)))
p_rarefaction100
```

### Rarefaction curves

One of the most informative application that involves rarefaction and microbial richness is the creation of rarefaction curves:

```{r, message=FALSE, warning=FALSE}
# devtools::install_github("gauravsk/ranacapa")
library(ranacapa)
set.seed(123)
rarefaction_curves <- ggrare(
    mia::makePhyloseqFromTreeSE(gp_genus), 
    step = 10000, plot = FALSE, parallel = TRUE)
```

```{r, fig.cap="Rarefaction curves."}
library(ggplot2)
# https://rdrr.io/github/LTLA/scater/src/R/plot_colours.R
total_colors <- scater:::.get_palette("tableau10medium")[1:9]
names(total_colors) <- levels(gp_genus$SampleType)

print(rarefaction_curves) + 
    geom_line(aes(color = SampleType)) + 
    facet_wrap(~ SampleType, nrow = 3, ncol = 3) + 
    scale_color_manual(values = total_colors) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
    labs(title = "Rarefaction curves",
         subtitle = "Faceted by sample type")
```

In this case the process of rarefaction consists in repeatedly subsampling the sequencing data at different depths, typically starting from the smallest number of reads in any sample to the largest. At each subsampling depth, the number of observed taxa is calculated (*i.e.*, richness), and these values are plotted against the corresponding sequencing depth. The resulting curve shows how the diversity of the microbial community changes as the sequencing depth increases.

The importance of rarefaction curves lies in several key aspects:

1.  **Assessing data completeness**. If the curves reach a plateau, it suggests that the sequencing depth is sufficient to capture most of the diversity in the sample. On the other hand, if the curves do not plateau (*e.g.*, one tongue sample, and all Sediment (estuary) samples), it indicates that more sequencing effort is needed to adequately characterize the microbial diversity.

2.  **Fair comparison**. Rarefaction curves enable fair comparisons of diversity across different samples with varying sequencing depths. By subsampling all samples to the same sequencing depth, the curves provide a standardized view of the community's diversity, facilitating reliable comparisons.

3.  **Assessing quality**. Rarefaction curves can help identify samples with low sequencing depth that might require additional attention or potentially be excluded from further analysis due to insufficient data.

4.  **Study design optimization**. Researchers can use rarefaction curves during the study design phase to estimate the required sequencing depth for reliable and comprehensive community profiling. This information can help optimize cost-effectiveness and resource allocation.

### Bray-Curtis dissimilarity index

Similarly to what we did for microbial richness, we can also study the effect of rarefaction on beta-diversity. In particular, we will focus on the Bray-Curtis dissimilarity index, a widely used metric to quantify the dissimilarity or similarity between two ecological samples, typically used in the context of community ecology and biodiversity studies.

Here we start by computing the Bray-Curtis indexes on raw counts, relative abundances (Total Sum Scaling normalization), and some subsamplings (the same used earlier for richness). Then use Multi Dimensional Scaling unsupervised ordination method is used to visualize the results.

```{r, warning=FALSE, message=FALSE, fig.width=10, fig.height=7}
gp_genus <- transformCounts(gp_genus, 
    assay.type = "counts", method = "relabundance")

gp_genus <- runMDS(gp_genus,
    FUN = vegan::vegdist,
    method = "bray",
    name = "PCoA_BC_raw",
    assay.type = "counts")

gp_genus <- runMDS(gp_genus,
    FUN = vegan::vegdist,
    method = "bray",
    name = "PCoA_BC_TSS",
    assay.type = "relabundance")

gp_genus <- runMDS(gp_genus,
    FUN = vegan::vegdist,
    method = "bray",
    name = "PCoA_BC1",
    assay.type = "rare1")

gp_genus <- runMDS(gp_genus,
    FUN = vegan::vegdist,
    method = "bray",
    name = "PCoA_BC2",
    assay.type = "rare2")

gp_genus <- runMDS(gp_genus,
    FUN = vegan::vegdist,
    method = "bray",
    name = "PCoA_BC3",
    assay.type = "rare3")

p_bc_raw <- plotReducedDim(gp_genus, "PCoA_BC_raw", 
    colour_by = "SampleType") + 
    labs(title = "Raw counts, Bray-Curtis, PCoA")
p_bc_tss <- plotReducedDim(gp_genus, "PCoA_BC_TSS", 
    colour_by = "SampleType") + 
    scale_y_reverse()  + 
    labs(title = "TSS, Bray-Curtis, PCoA")
p_bc_r1 <- plotReducedDim(gp_genus, "PCoA_BC1", 
    colour_by = "SampleType")  + 
    labs(title = "Subsampled seed = 1, Bray-Curtis, PCoA",
         subtitle = paste("N(L,min) =", min(gp_genus$sum)))
p_bc_r2 <- plotReducedDim(gp_genus, "PCoA_BC2", 
    colour_by = "SampleType") + 
    labs(title = "Subsampled seed = 2, Bray-Curtis, PCoA",
         subtitle = paste("N(L,min) =", min(gp_genus$sum)))
p_bc_r3 <- plotReducedDim(gp_genus, "PCoA_BC3", 
    colour_by = "SampleType") + 
    labs(title = "Subsampled seed = 3, Bray-Curtis, PCoA",
         subtitle = paste("N(L,min) =", min(gp_genus$sum)))

(p_bc_raw + p_bc_tss + plot_spacer()) / 
    (p_bc_r1 + p_bc_r2 + p_bc_r3) + 
    plot_layout(guides = "collect") +
    plot_annotation(tag_levels = "a")
```

The main difference in these panels is only between **a** and all the others which are quite similar. That's because in this case the total sum scaling normalization (panel **b**) behave similarly to using subsampling at the fixed library size of $N_{L,min}=58688$ (panels **c**, **d**, and **e**).

The Bray-Curtis dissimilarity between two samples is computed following the formula:

$$
BC_{jk} = \frac{\sum_i|x_{ij}-x_{ik}|}{\sum_ix_{ij}+x_{ik}}
$$

where $i$ is the number of taxa and $j,k$ are the sample indexes. By construction, when a sample has 10x more counts than the other, the numerator could be inflated just for this reason and not because of a real difference between the samples.

Practically, rarefaction consists in averaging the index measures of panels **b**, **c**, and **d** to obtain a single value for each sample.

```{r, message=FALSE, warning=FALSE}
# Recompute the distances
bc1 <- vegan::vegdist(x = t(assay(gp_genus, "rare1")), 
    method = "bray")
bc2 <- vegan::vegdist(x = t(assay(gp_genus, "rare2")), 
    method = "bray")
bc3 <- vegan::vegdist(x = t(assay(gp_genus, "rare3")), 
    method = "bray")

# Average the distances across subsamplings
bc_rarefaction <- (bc1 + bc2 + bc3) / 3

# Add the PCoA coordinates in a new reducedDim slot
# following the instructions in help("runMDS")
reducedDim(gp_genus, "PCoA_BCrarefaction") <- cmdscale(d = bc_rarefaction, k = 2, eig = TRUE)$points
```

```{r, warning=FALSE, message=FALSE}
plotReducedDim(gp_genus, "PCoA_BCrarefaction", 
    colour_by = "SampleType") + 
    labs(title = "Rarefaction Bray-Curtis, PCoA",
         subtitle = paste("subsamplings = 3, N(L,min) =", min(gp_genus$sum)))
```

As we did for the richness index, the process can be automatised:

```{r}
rarefaction_BC_PCoA <- function(
    tse, # The object to use
    assay.type = "counts", # Assay to use
    subsamplings = 10, # Number of subsamplings
    minLibSize = NULL, # The desired library size 
    seed = 123, # The seed
    BPPARAM = SerialParam()){ # Parallelisation
    
    counts <- assay(tse, "counts")
    
    if(is.null(minLibSize))
        minLibSize <- min(colSums(counts))
    
    dist_list <- bplapply(
        X = 1:subsamplings, 
        BPPARAM = BPPARAM, 
        BPOPTIONS = bpoptions(RNGseed = seed), 
        FUN = function(sub){
            # Subsampling step
            tse <- mia::subsampleCounts(tse, 
                assay.type = assay.type, 
                min_size = minLibSize,
                # seed = runif(1, 0, .Machine$integer.max), 
                replace = FALSE, 
                verbose = FALSE, 
                name = "sub")
            # BC calculation step
            BC_dist <- vegan::vegdist(
                x = t(assay(tse, "sub")),
                method = "bray")
            return(BC_dist)
    })
    # Compute the mean BC distance
    avg_BC_dist <- Reduce(f = '+', dist_list) / subsamplings
    # Compute the PCoA coordinates and return them
    cmdscale(d = avg_BC_dist, k = 2, eig = TRUE)$points
}
```

We run it in parallel (Linux or MacOS, Windows users should use `BPPARAM = SerialParam()`) and we directly store the computed values inside the `TreeSummarizedExperiment` object in the `reducedDim` slot.

```{r, warning=FALSE, message=FALSE}
reducedDim(gp_genus, "PCoA_BCrarefaction100") <- 
    rarefaction_BC_PCoA(
        tse = gp_genus, 
        assay.type = "counts", 
        subsamplings = 100,
        minLibSize = NULL, 
        seed = 321, 
        BPPARAM = MulticoreParam(4))
```

Finally, the graphical representation of rarefied beta-diversity using Bray-Curtis dissimilarity index and PCoA ordination (seed = 123, number of subsamplings = 100, $N_{L,min} = 58688$) is reported below.

```{r}
plotReducedDim(gp_genus, "PCoA_BCrarefaction100", 
    colour_by = "SampleType") + 
    labs(title = "Rarefaction Bray-Curtis, PCoA",
         subtitle = paste("subsamplings = 100, seed = 321, N(L,min) =", min(gp_genus$sum))) + 
    scale_y_reverse()
```

In this scenario rarefaction and Total Sum Scaling (TSS) normalization produced similar results. This is probably due to the fact that all samples are already at a sufficient depth to capture most of the microbial diversity.

But this can also occur when:

-   the differences in sequencing depth between samples are relatively small;

-   the samples have similar distributions of taxa and similar alpha diversity measures, so the impact of rarefaction or TSS normalization is minimal.

However, in cases where there are substantial differences in sequencing depth or if the samples have highly variable distributions of taxa, rarefaction and TSS normalization might lead to different results. In such situations, researchers should carefully consider which normalization approach is more appropriate for their specific dataset and research objectives.

*Presented at the [DataScience Group meeting](https://github.com/quadram-institute-bioscience/datasciencegroup/#readme), [Quadram Institute Bioscience](https://quadram.ac.uk/), July 2023*

## References
